---
title: "Market Basket Analysis"
output: html_document
date: "2024-12-16"
---

#1. Data Preparation for Market Basket Analysis

```{r}
mba <- read.csv("C:/Users/jivko/Documents/Data Analytics, Big Data, and Predictive Analytics/Personal Project/Market Basket Analysis/Groceries_dataset.csv")

# Print the first few rows
head(mba, 20)
```
```{r}
str(mba)
```
```{r}
install.packages("arules")
library(arules)
```

```{r}
# Combine 'Member_number' and 'Date' into a single identifier
mba_data <- data.frame(TransactionID = paste(mba$Member_number, mba$Date, sep = "_"),
                       Item = mba$itemDescription)

# Aggregate items by TransactionID
transactions <- split(mba_data$Item, mba_data$TransactionID)

# Convert to a transaction object
transactions <- as(transactions, "transactions")
```
```{r}
str(mba_data)
```
```{r}
head(mba_data, 10)
```

#2. Market Basket Analysis with 1 antecedent and 1 consequent 
```{r}
# Apply apriori to generate rules
rules <- apriori(transactions, 
                 parameter = list(supp = 0.0008, conf = 0.20))

rules <- rules[which(size(lhs(rules)) == 1 & size(rhs(rules)) == 1)]

# View a summary of the filtered rules
summary(rules)

# Inspect the top 10 filtered rules
inspect((rules))
```

```{r}
inspect(rules)
```

```{r}
# Load required packages
library(arules)
library(arulesViz)

# Visualize: Graph
plot(rules, method = "graph", engine = "htmlwidget")
```

#3. Market Basket Analysis with 2 antecedents and 1 consequent 
```{r}
# Rules with 2 items in lhs and 1 in rhs
rules_len2 <- apriori(transactions, 
                      parameter = list(supp = 0.0008, conf = 0.20))
rules_len2 <- rules_len2[which(size(lhs(rules_len2)) == 2 & size(rhs(rules_len2)) == 1)]

# View a summary of the filtered rules
summary(rules_len2)

inspect(head(rules_len2))
```

```{r}
inspect(rules_len2)
```

```{r}
# Load required packages
library(arules)
library(arulesViz)

# Visualize: Scatter Plot
plot(rules_len2, method = "scatterplot", measure = c("support", "confidence"), shading = "lift")

# Visualize: Grouped Matrix Plot
plot(rules_len2, method = "grouped")

# Visualize: Graph
plot(rules_len2, method = "graph", engine = "htmlwidget")
```

#4. Market Basket Analysis with 3 antecedents and 1 consequent 
```{r}
# Rules with 2 items in lhs and 1 in rhs
rules_len23 <- apriori(transactions, 
                      parameter = list(supp = 0.00032, conf = 0.20))
rules_len23 <- rules_len23[which(size(lhs(rules_len23)) == 3 & size(rhs(rules_len23)) == 1)]

# View a summary of the filtered rules
summary(rules_len23)

inspect(head(rules_len23))
```

```{r}
inspect(rules_len23)
```

```{r}
# Load required packages
library(arules)
library(arulesViz)

# Visualize: Scatter Plot
plot(rules_len23, method = "scatterplot", measure = c("support", "confidence"), shading = "lift")


# Visualize: Graph
plot(rules_len23, method = "graph", engine = "htmlwidget")
```

#5. Market Basket Analysis with 4 antecedents and 1 consequent 
```{r}
# Rules with 2 items in lhs and 1 in rhs
rules_len233 <- apriori(transactions, 
                      parameter = list(supp = 0.0002, conf = 0.20))
rules_len233 <- rules_len233[which(size(lhs(rules_len233)) == 4 & size(rhs(rules_len233)) == 1)]

# View a summary of the filtered rules
summary(rules_len233)

inspect(head(rules_len233))
```

```{r}
inspect(rules_len233)
```

```{r}
# Load required packages
library(arules)
library(arulesViz)

# Visualize: Scatter Plot
plot(rules_len233, method = "scatterplot", measure = c("support", "confidence"), shading = "lift")

# Visualize: Graph
plot(rules_len233, method = "graph", engine = "htmlwidget")
```

#6. Frequent Pattern Analysis
```{r}
# Load the arules package
library(arules)

# Count the frequency of each item in the transactions dataset
item_frequencies <- itemFrequency(transactions, type = "absolute")

# Sort the items by frequency and view the result
sorted_item_frequencies <- sort(item_frequencies, decreasing = TRUE)

# View the top items
head(sorted_item_frequencies, 5)
```

#7. K-Means Clustering of Market Basket Data: Identifying Item Frequency Groups

```{r}
library(dplyr)
library(ggplot2)

# Aggregate data
basket_summary <- mba_data %>%
    group_by(Item) %>%
    summarise(Frequency = n())

# Scale the data (this is important for distance-based methods like k-means)
basket_data <- scale(basket_summary$Frequency)

# Range of k values to try
k_values <- 1:10

# Store the WSS values
wss_values <- sapply(k_values, function(k) {
    kmeans(basket_data, centers = k, nstart = 25)$tot.withinss
})

# Plot the Elbow Method
ggplot(data.frame(k = k_values, WSS = wss_values), aes(x = k, y = WSS)) +
    geom_line() +
    geom_point() +
    labs(title = "Elbow Method for Optimal k",
         x = "Number of Clusters (k)",
         y = "Total Within-Cluster Sum of Squares (WSS)") +
    theme_minimal()
```


```{r}
# Apply K-Means with the optimal k (k = 3)
set.seed(123)
kmeans_result <- kmeans(basket_data, centers = 3)

# Add the cluster assignment to the basket_summary data frame
basket_summary$Cluster <- as.factor(kmeans_result$cluster)

# Scatterplot of frequency vs clusters
library(ggplot2)

ggplot(basket_summary, aes(x = Frequency, y = Cluster, color = Cluster)) +
    geom_point(size = 3) +  # Plot points
    labs(title = "Scatterplot of Frequency vs Clusters",
         x = "Frequency",
         y = "Cluster") +
    theme_minimal() +
    scale_color_brewer(palette = "Set1") +
    theme(legend.position = "bottom")
```

```{r}
# Add a new descriptive column for clusters with specific labels
basket_summary$Cluster_Label <- case_when(
  kmeans_result$cluster == 1 ~ "Low Frequency Items",
  kmeans_result$cluster == 2 ~ "Medium Frequency Items",
  kmeans_result$cluster == 3 ~ "High Frequency Items"
)

# View the updated data frame
head(basket_summary)
```













